Growth model analysis 3spp WB

### 2019-03-15 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
add full interactions in s()

```{r functions}

riverOrderedIn <- factor(c('west brook', 'wb jimmy', 'wb mitchell',"wb obear"),levels=c('west brook', 'wb jimmy', 'wb mitchell',"wb obear"),labels = c("WB","OL","OS","IL"), ordered = T)

@@

```{r Load data, message=FALSE}
library(itsadug)
library(arm)
library(relaimpo)
library(linkedModels)
library(arrayhelpers)
library(tidyverse)
library(gridExtra)

load("D:/projects/linkedModels/data/out/ddD_bktbntats_1997.RData")
load("D:/projects/linkedModels/data/out/ddddD_bktbntats_1997.RData")
load("D:/projects/linkedModels/data/out/rawCountsAndMasses.RData")

mcmc <- list()
dGIn <- list()
ddGIn2 <- list()

load("D:/projects/linkedModels/data/out/dG_1530971738_crossValFALSE_bkt1997_Nimble.RData")

# input data
d1 <- ddG[[1]][[1]]
d2 <- ddG[[1]][[2]]
meanSampleIntervalMean <- data.frame( meanSampleInterval = colMeans(d1$sampleIntervalMean), season = 1:d1$nSeasons )

#RECALCULATE gr for dataset. Filtering by maxSampleInterval in the creat ddddG step leaves in trailing lengths from the long samp intervals
d2 <- d2 %>%
  group_by(tag) %>%
  # arrange(tag,sampleNumber) %>%
  mutate( lagObservedWeight = lead(observedWeight),
          lagObservedLength = lead(observedLength),
          grWeight = exp(lagObservedWeight - observedWeight)/as.numeric((lagDetectionDate - detectionDate)),
          grLength = (lagObservedLength - observedLength)/as.numeric((lagDetectionDate - detectionDate))
  )  %>%
  ungroup()

#d2 %>% filter(tag=='00088cc059') %>% select(tag,riverOrdered,riverN,lagRiverN,sampleNumber,observedLength,lagObservedLength)

#input data frame
ddGIn2[['bkt']] <- ddG[[1]][[2]]
dGIn[['bkt']] <- dG
# output data
mcmc[['bkt']] <- mcmcProcessed

load("D:/projects/linkedModels/data/out/dG_1530971744_crossValFALSE_bnt1997_Nimble.RData")
ddGIn2[['bnt']] <- ddG[[1]][[2]]
dGIn[['bnt']] <- dG
mcmc[['bnt']]  <- mcmcProcessed

load("D:/projects/linkedModels/data/out/dG_1530993703_crossValFALSE_ats1997_Nimble.RData")
ddGIn2[['ats']] <- ddG[[1]][[2]]
dGIn[['ats']] <- dG
mcmc[['ats']]  <- mcmcProcessed

rm(ddG,dG,mcmcProcessed)

@@

# gam models
library(mgcv)
mod_bnt <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species == 'bnt'))
summary(mod1_bnt)

mod_bkt <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species == 'bkt'))
summary(mod_bkt)

mod_ats <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(species == 'ats'))
summary(mod_ats)

#######################
vB$yearFactor <- factor(vB$year)
vB$yearOrdered <- factor(vB$year, levels = 2000:2013, ordered = T)
vB$yearOrderedSpecies <- factor(paste(vB$yearOrdered,vB$species, sep = '_'), ordered = T)

vB$cohortFactor <- factor(vB$cohort)
vB$cohortOrdered <- factor(vB$cohort, levels = 2000:2013, ordered = T)
vB$cohortOrderedSpecies <- factor(paste(vB$cohortOrdered,vB$species, sep = '_'), ordered = T)

vB$speciesOrdered <- factor(vB$species, levels = c('bkt','bnt'), ordered = T)
vB$cohortSpeciesRiver <- factor(paste(vB$cohortOrdered,vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)

vB$speciesRiver <- factor(paste(vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)

@@

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohort)

mod6csr <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiver), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6csr)
plot(mod6csr)

predGrid <- expand.grid(cohortSpeciesRiver = unique(mod6csr$model$cohortSpeciesRiver),
                        ydayCumul = seq(250,1000, by = 30)) %>%
                 mutate(cohort = as.numeric(substr(cohortSpeciesRiver,1,4)),
                        species = substr(cohortSpeciesRiver,6,8),
                        river = substring(cohortSpeciesRiver,10))

pred <- predict(mod6csr, newdata = predGrid)#, type = 'lpmatrix')

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(0,300) +
  facet_grid(river ~ cohort)

mod6sr <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = speciesRiver), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6sr)

predGrid <- expand.grid(speciesRiver = unique(mod6sr$model$speciesRiver),
                        cohort = 2000:2013,
                        ydayCumul = seq(250,1000, by = 30)) %>%
                 mutate(species = substr(speciesRiver,1,3),
                        river = substring(speciesRiver,5))

pred <- predict(mod6sr, newdata = predGrid)#, type = 'lpmatrix')

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohort)

# try spline on year - no real advantage, except we can see the squiggle
mod6a <- gam(cf ~ species * riverOrdered + 
                              s(ydayCumul, bs = 'cr') + 
                              s(year), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), 
             method = "REML")
plot(mod6a, scale = 0)
summary(mod6a)

>>>>

@@

```{r Load data, message=FALSE}
library(itsadug)
library(arm)
library(relaimpo)
library(linkedModels)
library(arrayhelpers)
library(mgcv)
library(tidyverse)
library(gridExtra)

#RECALCULATE gr for dataset. Filtering by maxSampleInterval in the creat ddddG step leaves in trailing lengths from the long samp intervals
# d2 <- d2 %>%
#   group_by(tag) %>%
#   # arrange(tag,sampleNumber) %>%
#   mutate( lagObservedWeight = lead(observedWeight),
#           lagObservedLength = lead(observedLength),
#           grWeight = exp(lagObservedWeight - observedWeight)/as.numeric((lagDetectionDate - detectionDate)),
#           grLength = (lagObservedLength - observedLength)/as.numeric((lagDetectionDate - detectionDate))
#   )  %>%
#   ungroup()

@@

####################################
# gam models

cohortOffset <- 2006

# 1-way
vB$cohortFactor <- factor(vB$cohort)
vB$cohortOrdered <- factor(vB$cohort, levels = 2000:2013, ordered = T)
#vB$cohortFactorCentered <- vB$cohortFactor - 2006
vB$cohortCentered <- vB$cohort - cohortOffset

vB$speciesFactor <- factor(vB$species, levels = c('bkt','bnt'))
vB$speciesOrdered <- factor(vB$species, levels = c('bkt','bnt'), ordered = T)

vB$riverFactor <- factor(vB$river, levels = c('bkt','bnt'))
#vB$riverOrdered <- riverOrdered exists

# 2-way
vB$speciesRiverOrdered <- factor(paste(vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)
vB$speciesCohortOrdered <- factor(paste(vB$speciesOrdered,vB$cohortOrdered, sep = '_'), ordered = T)
vB$cohortRiverOrdered <- factor(paste(vB$cohortOrdered,vB$riverOrdered, sep = '_'), ordered = T)

# 3-way
vB$cohortSpeciesRiverOrdered <- factor(paste(vB$cohortOrdered,vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)

##########################################################
# by cohortSpeciesRiver - model is overfit?

modCSR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCSR)
#plot(modCSR)

predGridCSR <- expand.grid(cohortSpeciesRiver = unique(modCSR$model$cohortSpeciesRiver),
                           ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiver,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiver,6,8),
                           riverOrdered = substring(cohortSpeciesRiver,10),
                           cohortCentered = cohortOrdered - cohortOffset,
                           species = speciesOrdered,
                           river = riverOrdered)

predCSR <- predict(modCSR, newdata = predGridCSR)#, type = 'lpmatrix')

predGrid2CSR <- bind_cols(predGridCSR,as.data.frame(predCSR))

ggplot(predGrid2CSR, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,210) +
  facet_grid(river ~ cohort)

##########################################################
# by speciesRiver ##########################################################

modSR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modSR)
#plot(modSR, scale = 0)

predGridSR <- expand.grid(speciesRiverOrdered = unique(modSR$model$speciesRiverOrdered),
                          cohortOrdered = 2000:2013,
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesRiverOrdered,1,3),
                          riverOrdered = substring(speciesRiverOrdered,5),
                          cohortCentered = cohortOrdered - cohortOffset,
                          species = speciesOrdered,
                          river = riverOrdered)

predSR <- predict(modSR, newdata = predGridSR)#, type = 'lpmatrix')

predGrid2SR <- bind_cols(predGridSR,as.data.frame(predSR))

ggplot(predGrid2SR, aes(ydayCumul,predSR, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohortOrdered)

# by speciesCohort
system.time(
modSC <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
)
summary(modSC)
#plot(modSC, scale = 0)

predGridSC <- expand.grid(speciesCohortOrdered = unique(modSC$model$speciesCohortOrdered),
                          river = unique(vB$riverOrdered),
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesCohortOrdered,1,3),
                          cohortOrdered = as.numeric(substring(speciesCohortOrdered,5)),
                          cohortCentered = cohortOrdered - cohortOffset,
                          species = speciesOrdered)

predSC <- predict(modSC, newdata = predGridSC)#, type = 'lpmatrix')

predGrid2SC <- bind_cols(predGridSC,as.data.frame(predSC))

ggplot(predGrid2SC, aes(ydayCumul,predSC, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohortOrdered)

# by cohortRiver

modCR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered), 
               data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCR)
#plot(modCR, scale = 0)

predGridCR <- expand.grid(cohortRiverOrdered = unique(modCR$model$cohortRiverOrdered),
                          speciesOrdered = c('bkt','bnt'),
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(cohortOrdered = as.numeric(substr(cohortRiverOrdered,1,4)),
                          riverOrdered = (substring(cohortRiverOrdered,6)),
                          cohort = cohortOrdered,
                          cohortCentered = cohortOrdered - cohortOffset,
                          river = riverOrdered,
                          species = speciesOrdered)

predCR <- predict(modCR, newdata = predGridCR)#, type = 'lpmatrix')

predGrid2CR <- bind_cols(predGridCR, as.data.frame(predCR))

ggplot(predGrid2CR, aes(ydayCumul, predCR, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,200) +
  facet_grid(river ~ cohortOrdered)

#############################

# add year to smooths

modCSRY <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered) +
               s(year, bs = 'cr', by = speciesOrdered), 
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)

################################################################################

# base model

mod <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr'), 
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)

AIC(modCSRY,modCSR,modSR,modSC,modCR,mod)  %>% rownames_to_column() %>% arrange(AIC)

```{r early gams}
# some early exploration

@@

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohort)
```
